Expos\'e - Simulation of Nano Particles in a Laser Trap

### 2017-01-27 doc/expose.tex
much changes
achtung! geschwindigkeits-initialisierung geaendert!

\begin{document}
\title{Simulation of Nano Particles in a Laser Trap}
\author{Mathias H\"old, BSc.}
\date{2016}
\maketitle
\thispagestyle{empty}
\newpage
\section{Introduction}
%something about computer science in general, optical tweezers and the need to study the
%systems on computers yada yada

\newpage
\section{Motivation}
%introduction of the experiment, the problem and the idea
The starting point of this thesis is an experiment conducted by Gieseler et al \cite{Gieseler2014}. It is an optical tweezer experiment, where the
motion of a glass nanoparticle in a laser trap was used to investigate the fluctuation theorem\cite{Crooks1999}.
\subsection{Experimental setup}
In the experiment, a silica nano particle with a radius of about 75 nm and mass of about $3 \times 10^{-18}$kg is trapped in a laser beam within a
vacuum chamber. The trapping of the silica nano particle is achieved by a gradient force of the
laser beam acting on the particle. The experimental setup is depicted in Fig.~\ref{fig:setup}.\\
The particle fluctuates within the trap in all three spatial directions. These fluctuations can be approximated
such that they are decoupled, which means that they can be described by a one-dimensional Langevin equation:
\begin{equation}
    \label{eq:langevin}
    \ddot{x} + \Gamma_0 \dot{x} + \Omega^2_0x = \frac 1 m \left(F_\text{fluct} + F_\text{ext}\right)
\end{equation}
%The experiment is set up as follows: a glass sphere with radius $r \approx 75$ nm and mass $m \approx 3 \times 10^{-18}$ kg is trapped in a laser beam
%in a vacuum chamber. The initial state of the glass particle is prepared by modulation of the laser, such that the oscillation of the glass particle
%is suppressed in every direction. This is done by measuring the oscillation of the particle in every direction (assuming that the oscillation in every
%direction is decoupled) and using this information as feedback for the laser. This experimental setup can be seen in figure \ref{fig:setup}.
\begin{figure}[H]
    \begin{center}
        \includegraphics[scale=0.3]{images/experimental_setup.jpg}
        \caption{Experimental setup of the optical tweezer experiment. A silica nano particle is trapped in a laser beam via gradient force in a
        vacuum. The feedback is used to cool down the particle and create a non-equilibrium steady state. In the first part of the experiment, the
    feedback is turned off and the motion of the particle towards an equilibrated state is observed. In the second part of the experiment, the steady state
of the particle is modified by a parametric drive. Both the parametric drive and the feedback are turned off and -- as in the first part -- the
motion of the particle towards an equilibrated state is observerd. Figure taken from \cite{Gieseler2014}.}
        \label{fig:setup}
    \end{center}
\end{figure}
On the left hand side we have the position $x$ (and its derivatives $\dot{x}$, the velocity and $\ddot{x}$, the acceleration), the friction 
coefficient $\Gamma_0$ and the angular frequency $\Omega_0$ that describes the fluctuation along the chosen axis. On
the right hand side, there are two forces. The first one is $F_\text{fluct}$, which describes a stochastic force caused by interactions with the gas
in the vacuum chamber. This force is given by
\begin{equation}
    \label{eq:ffluct}
    F_\text{fluct} = \sqrt{2m\Gamma_0k_BT_0} \ \xi\left(t\right)
\end{equation}
where $T_0$ is the temperature of the heat bath (i.e., the surrounding gas in the vacuum chamber), $k_B$ is the Boltzmann constant and $\xi(t)$ is
white noise, which obeys the equations $\left\langle\xi(t)\right\rangle= 0$ and $\left\langle\xi(t)\xi(t')\right\rangle= \delta(t-t')$, which means
that it is a random force. The term $\Gamma_0$ appears in the formula \eqref{eq:ffluct} 
due to the fluctuation-dissipation theorem, which links the damping rate to the stochastic force.\\
%The external force $F_\text{ext}$ is part of the experimental setup, where the frequency of the fluctuation along an axis, $\Omega_0$, is measured and
%used to suppress the motion along said axis. This causes a decrease in the particle fluctuations and thus acts as a cooling mechanism for the particle
%in the trap. This process creates a non-equilibrium steady state $\rho_{ss}(u,\alpha)$, which is not known analytically. This state is the starting
%point of this thesis.\\
\subsection{Model of the Surrounding Gas}
J. Millen et al.~\cite{MillenJ.2014} worked on a similar experimental setup (without the feedback mechanism) and investigated the heating of the
particle in the trap as well as the gas surrounding the particle.\\
They start from the premise that there are 4 different temperatures one has to consider in such an experiment: the temperature of the gas particles
before interacting with the nano particle $T_\text{imp}$ (for \textit{impinging}), the temperature of the gas particles after 
interaction $T_\text{em}$ (for emerging), the surface temperature of the nano particle $T_\text{sur}$ and the temperature of the 
center of mass of the nano particle $T_\text{COM}$. The experimental setup
and the different temperatures are depicted in Fig.~\ref{fig:levitation}.\\
\begin{figure}[H]
    \begin{center}
        \includegraphics[scale=0.2]{images/nnano_millen.jpg}
        \caption{The experimental setup used by Millen et al. In \textbf{a} the experiment is schematically depicted and \textbf{b} shows a photograph
            of the actual experimental setup. In \textbf{c} the 4 different temperatures that are present during the experiment are depicted and
        \textbf{d} shows the position of the particle in the trap over time. The image was taken from~\cite{MillenJ.2014}}
        \label{fig:levitation}
    \end{center}
\end{figure}
Since the two temperatures of the gas, $T_\text{imp}$ and $T_\text{em}$, are not equal, the gas surrounding the nano particle is not in a thermal
equilibrium, which is in contrast to the assumption in many optical tweezer experiments. Since the gas particles only interact with the nano particle
in the trap and not wich each other, this situation creates two heat baths of different temperatures with the nano particle acting as a mediator
between those two baths. This situation is depicted in
Fig.~\ref{fig:heatbaths}.
\begin{figure}[H]
    \begin{center}
        \includegraphics[scale=0.5]{images/nano_nonequilibrium_cropped.jpg}
        \caption{Schematic picture of the two heat baths. The (hot) nano particle interacts with the (cold) incoming gas particles and they leave with
        a higher temperature. The incoming and outgoing particles do not interact with each other and thus create two separate heat baths with
    different temperatures. Image modified from \cite{Kroy2014}}
        \label{fig:heatbaths}
    \end{center}
\end{figure}
The one dimensional Langevin equation \eqref{eq:langevin} has to be modeled to include the different friction coefficients of the two heat baths. The
model derived in \cite{MillenJ.2014} is described in the following equation:
\begin{equation}
    M\ddot{x}(t) + M\left(\Gamma^\text{imp}+\Gamma^\text{em}\right)\dot{x}(t) + M\omega^2x(t) =F^\text{imp}+F^\text{em}
\end{equation}
where $x$ is the position of the nano particle, $\dot{x}$ and $\ddot{x}$ are the velocity and the acceleration, respectively, $M$ is the mass,
$\omega$ is the trap frequency, $\Gamma^\text{imp}$ and $\Gamma^\text{em}$ are the damping constants for the 
impinging particles and the emerging particles, respectively and $F^\text{imp}$ and $F^\text{em}$ are the associated noise terms, 
which satisfy $\langle F^\text{imp}\rangle= \langle F^\text{em} \rangle = 0$ individually.\\
Miller et al. investigate the dependence of the emerging temperature $T_\text{em}$ on the intensity of the laser for different sized nano particles
and different pressures.\\

%These four temperatures introduced in the paper, together with the surrounding pressure and the intensity of the laser 
%The existence of the two temperatures of the gas implies that, contrary to the assumption in many of the experiments, the gas is not in a thermal
%equilibrium. The Langevin equation (like the one in \eqref{eq:langevin}) has to be modeled k

\newpage
\section{Simulation}
%simulation techniques and general concept of the simulation itself, used methods and so on 
The problem at hand can be studied on an atomic level with the use of computer simulation. There is a variety of methods for computer simulations
that are widely used, one of which being molecular dynamics (MD) simulations.\\ 
The goal is to simulate the experiment depicted in Fig.~\ref{fig:setup} as accurately as possible. To achieve this, the setup has to be broken down
into individual pieced that can be modeled by using existing methods. The nano particle will be approximated by a cube of particles, sitting on an FCC
lattice that interact via a Lennard-Jones potential. The laser will be broken down into its two main purposes, as it acts as a trapping device as well
as a heat source for the nano particle. Surrounding the nano particle will be a gas chamber, that acts as a thermostat in the simulation, that will be
modeled as a box surrounding the nano particle. This box is filled with gas particles that interact with the nano particle via a soft-sphere potential
and do not interact with one another.\\
The following section will give a brief overview of the concepts used to simulate every part of the experiment.

\subsection{Molecular Dynamics}
Molecular dynamics~\cite{Frenkel2001} simulations is a technique for simulating, as the name suggests, the dynamics of a classical many-body system. In this case,
classical means, that the trajectories of the individual particles are calculated using classical mechanics rather then quantum mechanics. For
relatively big atoms/molecules this is a very good approximation, whereas for systems consisting of hydrogen or helium the effects of quantum
mechanics cannot be neglected and other methods have to be used.\\
The dynamics of the system are obtained by solving Newton's equations of motion for every particle.

\subsection{The Nano Particle}
The glass particle from the experiment will be modeled as a system of particles interacting via a Lennard-Jones pair potential, 
\begin{equation}
    \label{eq:lj}
    U(r) = 4\varepsilon\left[\left(\frac\sigma r\right)^{12} - \left(\frac\sigma r\right)^6\right]
\end{equation}
where $\varepsilon$ is the depth of the potential well (and thus its unit is energy) and $\sigma$ is the distance at which the potential is zero. 
The form of the potential and the relation to the parameters is depicted in Fig. \ref{fig:lj}.
\begin{figure}[h]
    \begin{center}
        \includegraphics[scale=0.2]{images/LJ-2.pdf}
        \caption{The Lennard-Jones 12-6 potential from \eqref{eq:lj}. The x-axis is the particle distance divided by $\sigma$ and  the y-axis is the
        potential divided by the depth of the potential well.}
        \label{fig:lj}
    \end{center}
\end{figure}
Since $\varepsilon$ and $\sigma$ are crucial parameters for the simulation and do not change over time, it is practical to use them to
define the dimensions of the system. This means that the unit of distance is $\sigma$, the unit of energy is $\varepsilon$ and the unit of mass is 
the mass of the simulated particle. The so called \textit{reduced units} can be constructed from these three parameters and put into relation to the
original units. Here are some examples:
\begin{itemize}
    \item {distance:} $r^* = r/\sigma$
    \item {potential energy:} $U^* = U/\varepsilon$
    \item {temperature:} $T^* = k_B T/\varepsilon$
    \item {time:} $t^* = t\sqrt{\varepsilon/(m\sigma^2)}$
    \item {pressure:} $P^* = P\sigma^3/\varepsilon$
    \item {density:} $\rho^* = \rho \sigma^3$
\end{itemize}
One very popular choice for the simulated atoms is Argon because it is an inert gas and the atoms behave 
approximately like hard spheres which attract each other with weak van der Waals forces, which justifies the use of the Lennard-Jones potential. 
Argon has a mass of m = $6.69 \times 10^{-26}$ kg, $\sigma = 3.4 \times 10^{-10}$m and $\varepsilon = 1.65 \times 10^{-21}$J.\\
With the above introduced reduced units, the Lennard-Jones potential can be written as
\begin{equation}
    U(r^*) = 4\left[{r^*}^{-12} - {r^*}^{-6}\right].
\end{equation}
Since the reduced units will be used throughout the rest of this thesis, I will drop the asterisk henceforth.\\
From the Lennard-Jones potential the corresponding force can be calculated by taking the derivative with respect to the direction of interest:
\begin{eqnarray}
    F_{x} &=& -\frac{\partial}{\partial x} U(r) \nonumber\\
                &=& -\frac{\partial}{\partial x} 4\left[{r}^{-12} - {r}^{-6}\right] \nonumber\\
                &=& -4 \left[(-12){r}^{-13} - (-6){r}^{-7}\right] \frac{\partial r}{\partial x} \nonumber\\
                &=& 48 \left[r^{-13} - 0.5 \ r^{-7}\right] \frac{x}{r} \nonumber\\
    \label{eq:ljforce} &=& 48 \left[r^{-14} - 0.5 \ r^{-8}\right] x
\end{eqnarray}
The force in the y and z direction can be calculated analogously.\\
The initial configuration of the particles is a face centered cubic (FCC) lattice. A schematic of the FCC lattice is depicted in Fig. \ref{fig:fcc}.
With the choice of FCC as initial configuration, there are optimal numbers for the numbers of the particles in the system. Since once FCC cell 
shares its atoms with its next neighbours, the number of atom per unit cell is 4 -- 1/8 of a particle on eight corners and 1/2 of a particle on six faces.
The whole system of atoms is then created by repeating this cell structure. One convenient way is to arrange the unit cells in a cubic system, 
so if there are $M$ FCC unit cells on one edge, the whole system consists of $M^3$ cells. Since there are 4 particles per cell, 
there are ideal or so called \textit{magic numbers} for atoms for which this setup works perfectly: $N = 4M^3 = 4,32,108,256,500,864,\ldots$.\\
\begin{figure}[h]
    \begin{center}
        \begin{subfigure}[t]{0.4\textwidth}
            \includegraphics[scale=0.3]{images/fcc-2.pdf}
            \caption{Schematic figure of a face centered cubic (FCC) lattice}
            \label{fig:fcc}
        \end{subfigure} 
        \
        %\hspace{60pt}
        \begin{subfigure}[t]{0.4\textwidth}
            \includegraphics[scale=0.3]{images/unit_cell.pdf}
            \caption{Setup for successively building a FCC lattice}
            \label{fig:unitcell}
        \end{subfigure}
    \end{center}
\end{figure}
%\begin{figure}
    %\begin{center}
        %\includegraphics[scale=0.3]{images/fcc_unit.png}
        %\caption{Schematic of the number of particles contained in one FCC unit cell. The atoms on the edges are shared by 8 cells and the ones on the
            %faces by 2 cells. Image taken from
            %{https://upload.wikimedia.org/wikipedia/commons/2/27/Elementarzelle\_einer\_kubisch\_raumzentrierten\_Elementarzelle.png}}
    %\end{center}
%\end{figure}
% do i need to credit the uni buffalo course? 
% http://www.physics.buffalo.edu/phy411-506/
There are several ways to achieve this initial configuration and the one used in this thesis \cite{buffalo} was to create a kind of unit cell 
consisting of four atoms, as shown in Fig. \ref{fig:unitcell}, which can be described by a set of points
\begin{eqnarray*}
    p_1 &=& \{0,0,0\}\\
    p_2 &=& \{0.5*a,0.5*a,0\}\\
    p_3 &=& \{0.5*a,0,0.5*a\}\\
    p_4 &=& \{0,0.5*a,0.5*a\}
\end{eqnarray*}
From the particle number $N$ and the number of FCC unit cells per edge $M$ the \textit{lattice constant} $a$ can be calculated
\begin{equation}
    a = \frac{L}{M}
\end{equation}
where $L$ is the side length of the cube that is the whole system and it is calculated via the density of the system
\begin{equation}
    L = \sqrt[3]{\frac{N}{\rho}}
\end{equation}
With the lattice constant and the 4 points of the FCC cell, all the particles can be put into place.

\subsection{The Velocity-Verlet Algorithm}
\label{section:verlet}
When we look at the system from a microscopic standpoint, we see that it follows some kind of path in the phase space as time progresses. Every
point in this space corresponds to a set of positions and momenta and the connection between two points corresponds to the evolution of the system
from one state to another. As mentioned above, this evolution (the dynamics of the system) is a crucial element to molecular dynamics. Since the
equations of motion cannot be solved analytically in general, we need to approximate the solution.\\
The method used here is called finite difference approach. The trajectory of the system in the phase space is cut into finite pieces of length $\Delta
t$ and the equations of motion are solved for every segment separately (see Fig. \ref{fig:finitedifference}).\\
\begin{figure}
    \begin{center}
        \includegraphics[scale=0.5]{images/finite_approach.pdf}
        \caption{Simplified graphical schematic of the finite difference approach. The evolution of the system from a point $(r^N(t),p^N(t))$ in the
            phase space is approximated by slicing it up into pieces of length $\Delta t$. On every stop after the starting point ($t+\Delta t$, 
            $t+2\Delta t$, $t+3\Delta t$,$\ldots$) the equations of motion can be solved numerically.}
        \label{fig:finitedifference} 
    \end{center}
\end{figure}
There are several ways to solve this kind of problem, but since we are interested in implementing it into a computer program the ideal solution
should have some basic properties \cite{Allen1989}:
\begin{itemize}
    \item It should be fast and require little memory
    \item It should permit the use of a large time step $\Delta t$
    \item It should duplicate the classical trajectory as closely as possible
    \item It should satisfy known conservation laws
    \item It should be simple and easy to program
\end{itemize}
One algorithm that has all of the above mentioned features is the one proposed by Verlet \cite{Verlet1967}. In his paper, Verlet starts by Taylor
expanding the coordinate vector $\mathbf{r}_i$ for one particle after one time step $\Delta t$:
\begin{equation}
    \label{eq:verlet}
    \mathbf{r}_i(t+\Delta t) = \mathbf{r}_i(t) + \dot{\mathbf{r}}_i \Delta t + \frac12 \ddot{\mathbf{r}}_i \Delta t^2 + \frac1{3!} \dddot{\mathbf{r}}_i \Delta t^3 +
    \mathcal{O}(\Delta t^4)
\end{equation}
Since the first derivative of the coordinate is the velocity, $\dot{\mathbf{r}}_i(t) = \mathbf{v}_i(t)$, and the second derivative of the coordinate 
is the acceleration, $\ddot{\mathbf{r}}_i(t) = \mathbf{a}_i(t)$, using Newton's law $\mathbf{F}_i(t) = m_i \mathbf{a}_i(t)$ equation \eqref{eq:verlet} 
can be written as:
\begin{equation}
    \label{eq:verletnew}
    \mathbf{r}_i(t+\Delta t) = \mathbf{r}_i(t) + {\mathbf{v}}_i(t) \Delta t + \frac1{2m} {\mathbf{F}}_i(t) \Delta t^2 + \frac1{3!} \dddot{\mathbf{r}}_i(t)
    \Delta t^3+\mathcal{O}(\Delta t^4)
\end{equation}
The same calculation can be carried out for one time step before $t$:
\begin{equation}
    \label{eq:verletnewminus}
    \mathbf{r}_i(t-\Delta t) = \mathbf{r}_i(t) - {\mathbf{v}}_i(t) \Delta t + \frac1{2m} {\mathbf{F}}_i(t) \Delta t^2 - \frac1{3!} \dddot{\mathbf{r}}_i(t) \Delta t^3+
    \mathcal{O}(\Delta t^4)
\end{equation}
The sum of those two equations yields
\begin{equation}
    \mathbf{r}_i(t+\Delta t) + \mathbf{r}_i(t-\Delta t) = 2\mathbf{r}_i(t) + \frac1{2} {\mathbf{F}}_i(t) \Delta t^2 + \mathcal{O}(\Delta t^4)
\end{equation}
and from this we get the final form for the new coordinates:
\begin{equation}
    \label{eq:verletfinal}
    \mathbf{r}_i(t+\Delta t) = 2\mathbf{r}_i(t)+ \mathbf{r}_i(t-\Delta t)+ \frac1{2} {\mathbf{F}}_i(t) \Delta t^2
\end{equation}
This means that the new coordinates can be calculated using the current coordinates, the current forces and the coordinates from the past time step.\\
From equation \eqref{eq:verletfinal} one thing becomes clear: the velocities are not necessary to calculate the new positions and they are not
computed in the process, but for the calculation of i.e. the kinetic energy the velocities are needed. Thus, the velocities have to be calculated with
a combination of the Taylor expansions of the coordinate vectors at $t+\Delta t$ and $t-\Delta t$:
\begin{equation}
    \mathbf{r}_i(t+\Delta t) - \mathbf{r}_i(t-\Delta t) = 2\mathbf{v}_i(t)\Delta t + \mathcal{O}(\Delta t^3)
\end{equation}
which can be rewritten as
\begin{equation}
    \label{eq:velocities}
    \mathbf{v}_i(t) = \frac{\mathbf{r}_i(t+\Delta t) - \mathbf{r}_i(t-\Delta t)}{2\Delta t} + \mathcal{O}(\Delta t^2)
\end{equation}
This approach has the advantage that it is fast, requires little memory and is reliable in the sense that there is no energy drift occuring during the
simulation, which means that the energy is conserved. When we compare this to the list of desired properties this seems like a good algorithm.\\
This algorithm however has two significant disatvantages. The first one is the calculation of the velocities. As can be seen in equation
\eqref{eq:velocities}, the accuracy of the calculation is only $\mathcal{O}(\Delta t^2)$ while the positions can be calculated with an error of order
$\Delta t^4$. The other big disatvantage is the first step of the algorithm. Since the calculation of the new positions requires the current positions
and the ones from one time step before, which technically do not exist.\\
The solutions to this problem is to include the stepwise calculation of the velocities \cite{swope1982}. For this we start again with the Taylor
expansion of the coordinates mathbftor \eqref{eq:verletnew} and instead of Taylor expanding $\mathbf{r}_i(t-\Delta t)$, we write it as 
\begin{equation}
    \mathbf{r}_i(t) = \mathbf{r}_i(t+\Delta t) - \mathbf{v}_i(t+\Delta t) \Delta t + \frac1{2m} \mathbf{F}_i(t+\Delta t) \Delta t^2
\end{equation}
Using \eqref{eq:verletnew} in the abvove equation we get
\begin{equation}
    \begin{aligned}
        \mathbf{r}_i(t) &= \mathbf{r}_i(t) + \mathbf{v}_i(t) \Delta t + \frac1{2m}\mathbf{F}_i(t)\Delta t^2 \\
                     &- \mathbf{v}_i(t+\Delta t) \Delta t + \frac1{2m} \mathbf{F}_i(t+\Delta t) \Delta t^2
    \end{aligned}
\end{equation}
and thus
\begin{equation}
    \label{eq:vvelocities}
    \mathbf{v}_i(t+\Delta t) = \mathbf{v}_i(t) + \frac1{2m} \left(\mathbf{F}_i(t) + \mathbf{F}_i(t+\Delta t)\right) \Delta t
\end{equation}
With the addition of this equation this algorithm is called the Velocity-Verlet algorithm, which is summarized in the following two equations:
\begin{equation}
    \label{eq:velocityverlet}
    \begin{aligned}
        \mathbf{r}_i(t+\Delta t) = \mathbf{r}_i(t) + {\mathbf{v}}_i(t) \Delta t + \frac1{2m} {\mathbf{F}}_i(t) \Delta t^2\\
        \mathbf{v}_i(t+\Delta t) = \mathbf{v}_i(t) + \frac1{2m} \Big[\mathbf{F}_i(t) + \mathbf{F}_i(t+\Delta t)\Big] \Delta t
    \end{aligned}
\end{equation}
This algorithm is self starting, uses a small amount of memory, conserves the energy (not exactly) 
and gives a very good approximation to the originial trajectory
in the phase space. To program this algorithm the following steps are needed: 
\begin{enumerate}
    \item Calculate all the forces between the particles (for the first step only)
    \item Calculate the new positions with current velocity and forces
    \item Calculate first half of the new velocities with current forces
    \item Calculate all the forces for the new position
    \item Use new forces to calculate second half of new velocities
\end{enumerate}
The first point only has to be carried out for the first step, because the forces have not been calculated at this point. As the forces are calculated
for the new positions in step 4, they can be used for the next time step. For the first time step the velocities have to be chosen randomly and are 
calculated with this algorithm from that point forward.

\subsection{The Laser Beam - Energy Influx}
The nano particle is trapped in the laser beam. While the motion of the center of mass is localized, the
individual atoms that make up the glass sphere absorb the energy from the laser which increases their velocity.\\
To simulate this kind of behaviour, thermostat algorithms \cite{Huenenberger2005} such as the Nos\'e-Hoover \cite{Nose1984,Hoover1985} or the Andersen
\cite{Andersen1980} algorithms are often used in simulation to change the temperature of the system in a controllable way.\\
The problem with these kind of algorithms is, that a target temperature has to be fixed which will be reached at some point. In order to simulate the
influx of energy from the laser it would be better to have an algorithm that supplies the system with a certain amount of energy continuously.
Fortunately, such an algorithm exists.\\
The algorithm is called Heat Exchange Algorithm (HEX) \cite{Hafskjold1993}. Its intended purpose is the use in non-equilibrium molecular dynamics
(NEMD) to study transport phenomena and determine transport coefficients. The algorithm works by introducing two regions in the system, one serving as
a heat source and the other as a heat sink. A specific amount of heat is then exchanged between those two reservoirs. As it turns out however, this
algorithm introduces an energy shift for longer simulation times. This led Wirnsberger et al. \cite{Wirnsberger2015} to revisit the algorithm and 
identify the cause of this energy drift, to create a more suitable algorithm, which they called eHEX.\\
As in the HEX algorithm, regions are introduced to the system which act either as heat sinks or heat sources. These are labelled with $\Gamma_k$,
where $k > 0$, and have corresponding amount of exchanged heat $\Delta Q_{\Gamma_k}$. If $\Delta Q_{\Gamma_k}$ is negative, heat is subtracted from
the system and vice versa. Regions that neither act as heat source or heat sink are labelled with $\Gamma_0$, which are also called
Hamiltonian regions (see Fig. \ref{fig:ehex}). 
\begin{figure}
    \begin{center}
        \includegraphics[scale=0.3]{images/ehex_system.png}
        \caption{Setup for the use of the HEX and eHEX algorithms. The simulationbox $\Omega$, which is moving with a velocity of $v_\Omega$,
        contains a heat source (red), $\Gamma_1$, which is moving with velocity $v_{\Gamma_1}$ and a heat sink (blue), $\Gamma_2$, which is moving 
        with velocity $v_{\Gamma_2}$. The regions that are neither heat sources or heat sinks are denoted by $\Gamma_0$.}
        \label{fig:ehex} 
    \end{center}
\end{figure}
The centers of mass of the particles in simulation box, denoted by $\Omega$, and the regions $\Gamma_k$ 
are assumed to be moving with velocities $v_\Omega$ and $v_{\Gamma_k}$ respectively.\\
The change of the energy in a region $\Gamma_k$ is achieved by rescaling the velocities by a factor $\xi_k$ and shifted by the velocity of the
corresponding region:
\begin{equation}
    \mathbf{v}_i \rightarrow \mathbf{\bar{v}}_i = \xi_k \mathbf{v}_i + (1-\xi_k)\mathbf{v}_{\Gamma_k}
\end{equation}
The bar over a quantity denotes the value after the exchange of heat.\\
The factor $\xi_k$ is given by
\begin{equation}
    \xi_k = \sqrt{1+\frac{\Delta Q_{\Gamma_k}}{\mathcal{K}_{\Gamma_k}}}
\end{equation}
where $\Delta Q_{\Gamma_k}$ is the exchanged heat in the region $\Gamma_k$ and $\mathcal{K}_{\Gamma_k}$ is the non-translational kinetic energy of the
region $\Gamma_k$ and is given by
\begin{equation}
    \label{eq:kineticK}
    \mathcal{K}_{\Gamma_k} = \sum_{i \in \gamma_k} \frac{m_i v_i^2}{2} - \frac{m_{\Gamma_k}v_{\Gamma_k}^2}{2}
\end{equation}
The sum is taken over all indices in $\gamma_k$ which is the set of indices of particles in the region $\Gamma_k$.\\
For the final version of the eHEX algorithm there are three more quantities needed. The first one is the heat flux per time step, denoted by 
$\mathcal{F}_\GammaK$:
\begin{equation}
    \mathcal{F}_\GammaK = \frac{\Delta Q_\GammaK}{\Delta t}
\end{equation}
The second one is the thermostatting force $\boldsymbol{\eta}_i$, which is defined as
\begin{equation}
    \boldsymbol{\eta}_i = 
    \begin{cases}
        m_i \ \dfrac{{\mathcal{F}_{\Gamma_{k(\mathbf{r}_i)}}}}{{2\mathcal{K}_{\Gamma_{k(\mathbf{r}_i)}}}} \
        \left(\mathbf{v}_i - \mathbf{v}_{\Gamma_{k(\mathbf{r}_i)}}\right)
        & \quad \text{if } k(\mathbf{r}_i) > 0 \\
        0 & \quad \text{otherwise}
    \end{cases}
    %\begin{cases}
        %m_i \ \frac{\displaystyle{\mathcal{F}_{\Gamma_{k(\mathbf{r}_i)}}}}{\displaystyle{2\mathcal{K}_{\Gamma_{k(\mathbf{r}_i)}}}} \
        %\left(\mathbf{v}_i - \mathbf{v}_{\Gamma_{k(\mathbf{r}_i)}}\right)
        %& \quad \text{if } k(\mathbf{r}_i) > 0 \\
        %0 & \quad \text{otherwise}
    %\end{cases}
\end{equation}
where $k(\mathbf{r}_i)$ is the index of the region in which particle $i$ located, i.e. $k(\mathbf{r}_i) = 0$ means that the particle is in a Hamiltonian
region and $k(\mathbf{r}_i) > 0$ denotes the heat sinks and sources.\\
The last quantity is the one that corrects the long term energy drift of the HEX algorithm, denoted by $\mathcal{E}{r}_{i,\alpha}$.
The analysis and derivation of this term is given in \cite{Wirnsberger2015}.
\begin{equation}
    \label{eq:bigeps}
    \begin{aligned}
        \mathcal{E}{r}_{i,\alpha} &= \dfrac{\eta_{i,\alpha}}{m_i\mathcal{K}_{\Gamma_{k(\mathbf{r}_i)}}} \left[
        \frac{\mathcal{F}_{\Gamma_{k(\mathbf{r}_i)}}}{48} + \frac{1}{6} \sum_{j\in\gamma_{k(\mathbf{r}_i)}} \mathbf{f}_j \cdot \left(\mathbf{v}_j - 
            \mathbf{v}_{\Gamma_{k(\mathbf{r}_i)}}\right)\right] \\
            &- \frac{\mathcal{F}_{\Gamma_{k(\mathbf{r}_i)}}}{12\mathcal{K}_{\Gamma_{k(\mathbf{r}_i)}}} \left[
            \frac{f_{i,\alpha}}{m_i} - \frac{1}{m_{\Gamma_{k(\mathbf{r}_i)}}} \sum_{j\in\gamma_{k(\mathbf{r}_i)}}f_{j,\alpha}\right]
    \end{aligned}
\end{equation}
The $\mathbf{f}$ in the above equation denotes the force corresponding to the chosen intermolecular potential $U(\mathbf{r})$, 
\begin{equation}
    \mathbf{f} = -\nabla_{\mathbf{r}_i} U(\mathbf{r}_i)
\end{equation}
With all the necessary quantities established, the updating sequence of the eHEX algorithm can be written down:
\begin{subequations}
\begin{eqnarray}
    \bar{\mathbf{v}}^n_i &=& \xi^n_{k(\mathbf{r}_i)} \mathbf{v}^n_i + \left(1-\xi^n_{k(\mathbf{r}_i)}\right) \mathbf{v}^n_{\Gamma_{k(\mathbf{r}_i)}} \\
    \bar{\mathbf{v}}^{n+\frac12}_i &=& \bar{\mathbf{v}}^n_i + \frac{\Delta t}{2m_i} \mathbf{f}^n_i\\
    \bar{\mathbf{r}}^{n+1}_i &=& \mathbf{r}^n_i + \Delta t \bar{\mathbf{v}}^{n+\frac12}_i\\
    \mathbf{f}^{n+1}_i &=& \left.-\nabla_{\mathbf{r}_i} U(\mathbf{r})\right|_{\mathbf{r} = \bar{\mathbf{r}}^{n+1}}\\
    \bar{\mathbf{v}}^{n+1}_i &=& \bar{\mathbf{v}}^{n+\frac12}_i + \frac{\Delta t}{2m_i} \mathbf{f}^{n+1}_i\\
    \mathbf{v}^{n+1}_i &=& \bar{\xi}^{n+1}_{k(\bar{\mathbf{r}}_i)} \bar{\mathbf{v}}^{n+1}_i + \left(1-\bar{\xi}^{n+1}_{k(\bar{\mathbf{r}}_i)}\right)

>>>>

\begin{document}
\title{Expos\'e - Simulation of Nano Particles in a Laser Trap}
\author{Mathias H\"old, BSc.}
\maketitle
\thispagestyle{empty}
\section{Konzept}
In dieser Masterarbeit wird mit Hilfe von Methoden der Computational Physics ein Experiment simuliert, in dem Nanopartikel in einer Laserfalle
gefangen werden. Das Hauptaugenmerk dabei liegt auf der Interaktion des Nanopartikels mit dem umgebenden Gas und dem Einfluss des Lasers auf die
Bewegung des Nanopartikels in der Falle. Die benutzte Technik in dieser Simulation nennt sich ,,molecular dynamics'' 
und wird dazu verwendet, um die zeitliche Entwicklung des Zustandes des Nanopartikels und des Gases zu berechnen.\\
Das Experiment besteht aus einem Nanopartikel, in diesem Fall eine Siliziumdioxid-Kugel mit einem Durchmesser von $r \approx 75 nm$, das in von einem
Laser durch Gradientenkraft lokalisiert wird. Das Nanopartikel befindet sich in einer Vakuumkammer mit variablem Druck.\\
Die Gasteilchen, die das Nanopartikel umgeben, interagieren nicht miteinander, wenn die mittlere freie Wegl\"ange der Gasteilchen viel gr\"o\ss er ist
als der Radius des Nanopartikels. Das bedeutet, dass die Interaktion nur zwischen den Gas Teilchen und dem Nanopartikel in der Falle stattfindet. 
Dadurch entstehen zwei W\"armereservoirs des umgebenden -- ein ,,kaltes'', bestehend aus Teilchen vor der Interaktion mit dem Nanopartikel und ein
,,hei\ss es'', bestehend aus Teilchen nach der Interaktion mit dem Nanopartikel.\\
Dem Nanopartikel selbst sind auch zwei Temperaturen zuzuweisen. Da es sich in einer Laserfalle befindet, absorbiert das Nanopartikel die Energie des
Lasers in Form von W\"arme, die an die innere Energie der Atome, aus dem das Nanopartikel besteht, abgegeben wird. Die Temperatur, die aus dieser
inneren Energie abzuleiten ist, wird ,,Oberfl\"achentemperatur'' genannt. Die zweite Temperatur, die dem Nanopartikel zuzuordnen ist, ist seine
Schwerpunktstemperatur -- jene Temperatur, die aus der Geschwindigkeit der Schwerpunktsbewegung berechnet wird. Durch die Lokalisierung in der
Laserfalle ist wird diese Bewegung minimiert. \\
Die Simulation dieses Experiments wird durch Modellierung der Einzelnen Bestandteile durchgef\"uhrt. Das Nanopartikel wird durch einen FCC Kristall
approximiert, in dem die Teilchen miteinander \"uber ein Lennard-Jones Potential miteinander wechselwirken. Der Laser wird durch zwei unterschiedliche
Teile modelliert. Zum einen wird die durch den Laser abgegebene und das Nanopartikel aufgenommene Energie durch den so genannten ,,eHEX'' -- enhanced
heat exchange algorithm -- modelliert. Die Lokalisierung des Nanopartikels durch den Laser wird durch ein harmonisches Potential modelliert, das auf
den Schwerpunkt wirkt. Das umgebende Gas wird durch einen Barostat modelliert, der in diesem Fall eher eine Funktion als Thermostat einnimmt. Die
Bewegung der einzelnen Atome wird durch den velocity Verlet Algorithmus berechnet.\\
Die Fragestellung, der in dieser Arbeit nachgegangen wird ist: Hat die erh\"ohte Temperatur der inneren Freiheitsgrade des Nanopartikels einen
Einfluss auf dessen Schwerpunktsbewegung?

\section{Vorl\"aufiges Inhaltsverzeichnis}
Die Arbeit wird in folgende Kapitel unterteilt sein:
\begin{itemize}
    \item{Introduction}
    \item{Motivation}
    \item{Simulation}
    \item{Results}
    \item{Conclusion}
\end{itemize}

\section{Auswahlbibliographie}
\nocite{*} 
\bibliography{references}
\bibliographystyle{unsrt}
